---
title: "Type-checker trends across CI runs"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    fig_width: 10
    fig_height: 6
params:
  workflow: "ci.yml"
  limit: 200
  cache_dir: "work/artifacts"
  top_n_packages: 24
  top_n_categories: 8
  upstream_repo: "programLyrique/r-c-typing"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(jsonlite)
  library(scales)
  library(DT)
})
theme_set(theme_minimal(base_size = 11))

cache_dir <- params$cache_dir
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
```

This report pulls every successful CI run's `results/` artifact via `gh`, caches it under
`` `r cache_dir` ``, and plots a time series for each run.

Render with:

```r
rmarkdown::render("scripts/trends.Rmd",
                  output_file = "../results/trends.html")
```

## Fetch artifacts

```{r fetch}
gh_path <- Sys.which("gh")
if (!nzchar(gh_path)) {
  stop("`gh` CLI is required. Install from https://cli.github.com/ and run `gh auth login`.")
}

runs_raw <- system2(
  "gh",
  c("run", "list",
    "--workflow", params$workflow,
    "--status", "success",
    "--limit", as.character(params$limit),
    "--json", "databaseId,number,createdAt,headSha,headBranch,event,displayTitle"),
  stdout = TRUE, stderr = TRUE
)
runs <- fromJSON(paste(runs_raw, collapse = "\n"))
runs$ts <- as.POSIXct(runs$createdAt, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
runs <- runs[order(runs$ts), , drop = FALSE]

cat(sprintf("Found %d successful runs (newest %s)\n",
            nrow(runs), format(max(runs$ts))))

# Download each artifact into cache_dir/<run_id>/. Skip if already present.
# `gh run download` exits non-zero if the artifact has expired or was renamed —
# we just drop the partial dir and continue.
for (i in seq_len(nrow(runs))) {
  rid <- as.character(runs$databaseId[i])
  rno <- as.character(runs$number[i])
  dest <- file.path(cache_dir, rid)
  if (file.exists(file.path(dest, ".downloaded"))) next
  unlink(dest, recursive = TRUE)
  dir.create(dest, recursive = TRUE, showWarnings = FALSE)
  status <- suppressWarnings(system2(
    "gh",
    c("run", "download", rid,
      "-n", paste0("results-", rno),
      "-D", dest),
    stdout = FALSE, stderr = FALSE
  ))
  if (status == 0L) {
    file.create(file.path(dest, ".downloaded"))
  } else {
    message(sprintf("run %s: artifact unavailable (skipped)", rid))
    unlink(dest, recursive = TRUE)
  }
}
```

```{r load}
# upload-artifact@v4+ may extract files at the root of the artifact or under
# `results/` depending on the action version that produced it. Look in both.
find_csv <- function(base, name) {
  candidates <- c(file.path(base, name), file.path(base, "results", name))
  hit <- candidates[file.exists(candidates)]
  if (length(hit) == 0L) NA_character_ else hit[[1]]
}

read_run <- function(rid, ts, run_no, sha, branch, title, event) {
  base <- file.path(cache_dir, as.character(rid))
  s_path <- find_csv(base, "summary.csv")
  if (is.na(s_path)) return(NULL)
  s <- read.csv(s_path, stringsAsFactors = FALSE)
  if (nrow(s) == 0L) return(NULL)
  s$run_id <- rid; s$run_no <- run_no; s$ts <- ts
  s$sha <- sha; s$branch <- branch
  s$commit_title <- title; s$event <- event

  f_path <- find_csv(base, "functions.csv")
  f <- if (!is.na(f_path)) {
    fd <- read.csv(f_path, stringsAsFactors = FALSE)
    if (nrow(fd) == 0L) {
      NULL
    } else {
      fd$run_id <- rid; fd$run_no <- run_no; fd$ts <- ts
      fd
    }
  } else NULL
  list(summary = s, functions = f)
}

loaded <- Map(read_run,
              runs$databaseId, runs$ts, runs$number,
              runs$headSha, runs$headBranch, runs$displayTitle, runs$event)
loaded <- Filter(Negate(is.null), loaded)

if (length(loaded) == 0L) {
  stop("No artifacts downloaded — the workflow's artifacts may have expired ",
       "(GitHub default retention is 90 days).")
}

summary_all   <- bind_rows(lapply(loaded, `[[`, "summary"))
functions_all <- bind_rows(lapply(loaded, `[[`, "functions"))

# Older artifacts predate some columns (e.g. n_timeout). Backfill with 0 so
# downstream code doesn't have to special-case missing columns.
for (col in c("n_timeout", "n_untypeable", "n_ep_typed", "n_typed", "n_functions")) {
  if (!col %in% names(summary_all)) summary_all[[col]] <- 0L
  summary_all[[col]][is.na(summary_all[[col]])] <- 0L
}

# Runs triggered by repository_dispatch (event-type r-c-typing-updated, fired
# from the sibling checker repo) carry "r-c-typing-updated" as displayTitle.
# Resolve them to the upstream commit subject by matching the run's start time
# to the most recent main-branch commit on `upstream_repo` at that point.
upstream_commits_path <- file.path(cache_dir, "upstream_commits.json")
fetch_upstream_commits <- function() {
  raw <- suppressWarnings(system2(
    "gh", c("api", "--paginate",
            paste0("repos/", params$upstream_repo,
                   "/commits?sha=main&per_page=100")),
    stdout = TRUE, stderr = FALSE
  ))
  if (!length(raw)) return(NULL)
  # --paginate concatenates multiple JSON arrays; jsonlite handles a single
  # array, so re-stitch the pages with simple bracket-balancing.
  txt <- paste(raw, collapse = "\n")
  txt <- gsub("\\]\\s*\\[", ",", txt)
  fromJSON(txt, simplifyDataFrame = FALSE)
}
if (file.exists(upstream_commits_path) &&
    difftime(Sys.time(), file.info(upstream_commits_path)$mtime, units = "hours") < 6) {
  upstream_commits <- fromJSON(upstream_commits_path, simplifyDataFrame = FALSE)
} else {
  upstream_commits <- tryCatch(fetch_upstream_commits(), error = function(e) NULL)
  if (!is.null(upstream_commits)) {
    write(jsonlite::toJSON(upstream_commits, auto_unbox = TRUE),
          upstream_commits_path)
  }
}

upstream_subject <- function(at_ts) {
  if (is.null(upstream_commits) || !length(upstream_commits)) return(NA_character_)
  for (c in upstream_commits) {  # commits are returned newest-first
    cd <- as.POSIXct(c$commit$committer$date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
    if (!is.na(cd) && cd <= at_ts) {
      msg <- c$commit$message
      return(strsplit(msg, "\n", fixed = TRUE)[[1]][1])
    }
  }
  NA_character_
}

is_dispatch <- summary_all$event == "repository_dispatch"
if (any(is_dispatch)) {
  resolved <- vapply(summary_all$ts[is_dispatch], upstream_subject, character(1))
  # Keep the original title where the lookup failed (e.g. upstream API down,
  # commit older than the fetch window).
  keep <- !is.na(resolved) & nzchar(resolved)
  summary_all$commit_title[is_dispatch][keep] <- resolved[keep]
}

# Per-run entry-point totals (sum across all native conventions).
summary_all <- summary_all %>%
  mutate(
    n_ep_total = coalesce(ep_call, 0L) + coalesce(ep_c, 0L) +
                 coalesce(ep_fortran, 0L) + coalesce(ep_external, 0L)
  )

cat(sprintf("Loaded %d runs, %d package-rows, %d function-rows\n",
            length(loaded), nrow(summary_all), nrow(functions_all)))
```

## Overall: typed vs. analyzed

```{r overall-typed}
totals <- summary_all %>%
  group_by(ts, run_no) %>%
  summarise(
    n_typed = sum(n_typed, na.rm = TRUE),
    n_functions = sum(n_functions, na.rm = TRUE),
    n_ep_typed = sum(n_ep_typed, na.rm = TRUE),
    n_ep_total = sum(n_ep_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_typed = ifelse(n_functions > 0, 100 * n_typed / n_functions, NA_real_),
    pct_ep    = ifelse(n_ep_total > 0, 100 * n_ep_typed / n_ep_total, NA_real_)
  )

ggplot(totals, aes(ts)) +
  geom_line(aes(y = n_functions, color = "analyzed"), linewidth = 0.6) +
  geom_line(aes(y = n_typed,     color = "typed"),    linewidth = 0.6) +
  geom_point(aes(y = n_functions, color = "analyzed"), size = 1.5) +
  geom_point(aes(y = n_typed,     color = "typed"),    size = 1.5) +
  labs(x = NULL, y = "Functions", color = NULL,
       title = "Typed vs. analyzed (totals over all packages)")
```

```{r overall-pct}
ggplot(totals, aes(ts, pct_typed)) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% typed", title = "% functions typed")
```

## Overall: entry-point typing rate

```{r overall-ep}
ggplot(totals, aes(ts, pct_ep)) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% entry points typed",
       title = "Entry-point typing rate (.Call/.C/.External/Fortran)")
```

## Run-by-run summary

One row per CI run: commit subject, total analyzed functions, % typed, and a
count column for each top error category.

```{r run-table}
run_totals <- summary_all %>%
  group_by(ts, run_no, sha, branch, commit_title) %>%
  summarise(
    n_analyzed = sum(n_functions, na.rm = TRUE),
    n_typed    = sum(n_typed,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_typed = ifelse(n_analyzed > 0, 100 * n_typed / n_analyzed, NA_real_)
  )

# Errors per category, per run. Use the same top-N selection logic as the
# error-trend section below so the columns match what's plotted.
errs_for_table <- functions_all %>%
  filter(!is.na(error_title), nzchar(error_title), status != "typed")
top_cats_tbl <- errs_for_table %>%
  count(error_title, sort = TRUE) %>%
  slice_head(n = params$top_n_categories) %>%
  pull(error_title)

errs_wide <- errs_for_table %>%
  mutate(error_title = ifelse(error_title %in% top_cats_tbl, error_title, "(other)")) %>%
  count(run_no, error_title, name = "n") %>%
  pivot_wider(names_from = error_title, values_from = n, values_fill = 0)

run_table <- run_totals %>%
  left_join(errs_wide, by = "run_no") %>%
  arrange(desc(ts)) %>%
  mutate(
    date      = format(ts, "%Y-%m-%d %H:%M"),
    pct_typed = round(pct_typed, 1)
  ) %>%
  select(n_analyzed, pct_typed, commit_title,
         run_no, date, branch,
         any_of(c(top_cats_tbl, "(other)")))

# Replace NA error counts (runs that predate a category or had zero of it)
# with 0 — a blank cell is harder to read than an explicit zero.
err_cols <- intersect(names(run_table), c(top_cats_tbl, "(other)"))
for (col in err_cols) run_table[[col]][is.na(run_table[[col]])] <- 0L

datatable(
  run_table,
  rownames   = FALSE,
  filter     = "top",
  extensions = "FixedHeader",
  options    = list(
    pageLength  = 25,
    order       = list(list(4, "desc")),  # 0-indexed: column 4 = date
    fixedHeader = TRUE,
    scrollX     = TRUE
  ),
  colnames   = c("# analyzed" = "n_analyzed",
                 "% typed"    = "pct_typed",
                 "commit"     = "commit_title",
                 "run"        = "run_no")
)
```

## Per-package: typed functions

Top `r params$top_n_packages` packages by analyzed-function count in the latest run.

```{r per-pkg-select}
latest_ts <- max(summary_all$ts)
top_pkgs <- summary_all %>%
  filter(ts == latest_ts) %>%
  arrange(desc(n_functions)) %>%
  slice_head(n = params$top_n_packages) %>%
  pull(package)
pkg_view <- summary_all %>% filter(package %in% top_pkgs)
```

```{r per-pkg-typed, fig.height=10}
ggplot(pkg_view, aes(ts)) +
  geom_line(aes(y = n_functions, color = "analyzed"), linewidth = 0.4) +
  geom_line(aes(y = n_typed,     color = "typed"),    linewidth = 0.4) +
  facet_wrap(~ package, scales = "free_y") +
  labs(x = NULL, y = "Functions", color = NULL,
       title = "Typed vs. analyzed per package")
```

```{r per-pkg-pct, fig.height=10}
pkg_view %>%
  mutate(pct_typed = ifelse(n_functions > 0, 100 * n_typed / n_functions, NA_real_)) %>%
  ggplot(aes(ts, pct_typed)) +
  geom_line(linewidth = 0.4) + geom_point(size = 0.6) +
  facet_wrap(~ package) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
  labs(x = NULL, y = "% typed", title = "% functions typed per package")
```

## Per-package: entry-point typing rate

Restricted to packages that have at least one entry point in the latest run.

```{r per-pkg-ep, fig.height=10}
ep_pkgs <- summary_all %>%
  filter(ts == latest_ts, n_ep_total > 0) %>%
  arrange(desc(n_ep_total)) %>%
  slice_head(n = params$top_n_packages) %>%
  pull(package)

summary_all %>%
  filter(package %in% ep_pkgs) %>%
  mutate(pct_ep = ifelse(n_ep_total > 0, 100 * n_ep_typed / n_ep_total, NA_real_)) %>%
  ggplot(aes(ts, pct_ep)) +
  geom_line(linewidth = 0.4) + geom_point(size = 0.6) +
  facet_wrap(~ package) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
  labs(x = NULL, y = "% entry points typed",
       title = "% entry points typed per package")
```

## Error categories: overall trend

`error_title` from `functions.csv`, restricted to non-typed rows. Top `r params$top_n_categories`
categories by total occurrence across all runs.

```{r errs-prep}
errs <- functions_all %>%
  filter(!is.na(error_title), nzchar(error_title), status != "typed")

top_cats <- errs %>%
  count(error_title, sort = TRUE) %>%
  slice_head(n = params$top_n_categories) %>%
  pull(error_title)

errs_top <- errs %>%
  mutate(error_title = ifelse(error_title %in% top_cats, error_title, "(other)"))

errs_overall <- errs_top %>%
  count(ts, run_no, error_title, name = "n")
```

```{r errs-overall}
ggplot(errs_overall, aes(ts, n, color = error_title)) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.2) +
  labs(x = NULL, y = "Functions", color = "Category",
       title = "Error categories over time (totals)")
```

```{r errs-overall-facet, fig.height=8}
ggplot(errs_overall, aes(ts, n)) +
  geom_line(linewidth = 0.5) + geom_point(size = 0.8) +
  facet_wrap(~ error_title, scales = "free_y") +
  labs(x = NULL, y = "Functions",
       title = "Error categories over time (faceted)")
```

## Error categories: per package

```{r errs-per-pkg, fig.height=12}
errs_pkg <- errs_top %>%
  filter(package %in% top_pkgs) %>%
  count(ts, package, error_title, name = "n")

ggplot(errs_pkg, aes(ts, n, color = error_title)) +
  geom_line(linewidth = 0.4) + geom_point(size = 0.6) +
  facet_wrap(~ package, scales = "free_y") +
  labs(x = NULL, y = "Functions", color = "Category",
       title = "Error categories per package")
```
